Gaps in stellar streams as a result of globular cluster fly-bys

### 2024-12-19 paper/main.tex
alessandra's 3rd round of comments

\institute{
  Paris Observatory. Paris Sciences et Lettres
  \email{salvatore.ferrone@obspm.fr}
  \and
    Dipartimento di Fisica, Universit\`a di Roma ``La Sapienza'',Piazza Aldo Moro\\
  \and
    Dipartimento di Fisica e Astronomia. ``Galileo Gallilei'' Università di Padova, Vicolo dell'Osservatorio 3 Padova 35122, Italy.
    \and
      Universit\'e de Strasbourg, CNRS, Observatoire astronomique de Strasbourg, UMR 7550, F-67000 Strasbourg, France
    \and
      Leibniz-Institut für Astrophysik Potsdam (AIP), An der Sternwarte 16, 14482 Potsdam, Germany
    \and
      Exeter
    % \thanks{}
  }

@@

  One of the interests in studying stellar streams is that they can constrain the gravitational field of their host galaxies, particularly the Milky Way. Compared to the measurement of the HI rotation curve, stellar streams offer the opportunity to investigate the potential of the host galaxy over a wide range of distances, reaching the outermost regions of the halo. For example, \citet{2011MNRAS.417..198V} demonstrated how stellar streams can infer the mass and scale parameters of dark matter halos using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. \citet{2018ApJ...867..101B} reviewed this concept from an information-theoretic point of view, identifying which orbits and configurations of stellar streams yield the most information about the galactic potential. However, the use of single streams for constraining the potential led to some ambiguous and not converging results over time. For example, \citet{2010ApJ...718.1128L} made use of the Sagittarius stream to infer that the dark matter halo of our Galaxy has a triaxial shape \citep[but see also][]{2004MNRAS.351..643H, 2005ApJ...619..800J, 2005ApJ...619..807L}; \citet{2016ApJ...833...31B} concluded that the dark matter halo of our Galaxy is nearly spherical at the distances of the Palomar~5 and GD-1 streams. The two contrasting results could also show that at one distance is the halo is triaxial and at another spherical showcasing the need for streams at different distances to map the halo shape. Recently, \citet{2024ApJ...967...89I} constructed the most accurate model of the Milky Way to date by applying a Monte Carlo Markov Chain (MCMC) fitting procedure. This method identified the set of potential parameters in an axisymmetric model of the Milky Way that best reproduces all observed stellar streams. This work represents a seminal achievement and a significant milestone in the field.

@@

\section{Methods}
  The most accurate model of the Palomar~5 stream would involve full modeling of the internal dynamics of the cluster. This would mean computing $N$-Body interactions with a $\mathcal{O}(N^2)$ computation time, stellar evolution, super novae, an initial mass distirbution, treatment of binary star systems, etc. \citep[for such an example][see,]{2021NatAs...5..957G}. Instead, We opt for solving the restricted-three body problem or particle-test method as we did for \citet{2023A&A...673A..44F}, which we present here for completeness.

@@

    Lastly, we select an integration time of 5~Gyr as a compromise between maximizing interaction statistics and adhering to the assumption that the galaxy can be modeled as a time-independent, constant mass distribution. \citet{2023A&A...673A.152I} analyzed the orbits of the Galactic cluster population using the same initial conditions as in this work, within five live Milky Way-like potentials from IllustrisTNG \citep{2018MNRAS.473.4077P}. They found that in all sampled potentials, orbital changes remain minimal over 5 Gyr, becoming significant only at earlier look-back times when the host galaxy had significantly less mass or was undergoing a merger event.

    \subsubsection*{ \texttt{Full} simulations}
    There is a primary methodological departure from \citet{2023A&A...673A..44F}.  In that paper, globular clusters evolved under the gravitational effect of the Galaxy only, while now we include also the effect of all the other Galactic globular clusters, by taking into account the direct $N$-body interactions among them. To quantify these interactions, all clusters are represented by Plummer spheres, each with mass and half-mass radius as reported in the Baumgardt catalog \citep{2021MNRAS.505.5957B}. For the remainder of this paper, we refer to these simulations, where the globular clusters interactions are taken into account as \texttt{full} simulations. \\
    For these simulations, we proceed into two steps:
      \begin{enumerate}
          \item First, starting from the Galactocentric positions and velocities of all 165 Galactic globular clusters, we integrate their orbits back in time for 5~Gyr under the influence of the Galaxy itself and their mutual influence. In the backward integration, the system of equations of motion for the globular clusters are thus: 
          \begin{equation}
          \ddot{\vec{r}}_i = -\nabla \Phi(R_i,z_i) + \left.\sum_{j\neq i}^{N_{GC}} \frac{Gm_j}{\left(|\vec{r}_j - \vec{r}_i|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j - \vec{r}_i\right),
          \end{equation}\label{eq:GCNBody} where $\vec{r}$ indicates the galactocentric position vector, the index $i$ indicates the globular cluster of interest with $R_i,z_i$ being its galactocentric positions in cylindrical coordinates; the index $j$ indicates the other globular clusters that are summed over. $N_{GC}$ is the total number of globular clusters which in this study is 165, $m_j$ is the mass of the j-th cluster in the sample, $b_j$ is its Plummer scale radius, and $\vec{r_j}$ is its galactocentric position. $\Phi$ represents the Galactic smooth potential \citep[][Model~II, in the present case]{2017A&A...598A..66P}. Note that the masses and sizes of the globular clusters are kept constant in these simulations and not allowed to vary with time. This implies that we do not take into account in these models the internal evolution of the globular clusters, nor their response to the Galactic tidal field.

        \item Once the positions and velocities of the entire globular cluster system, as to 5~Gyr ago, are found, we sample Palomar~5 with  100,000 particles which follow a Plummer distribution of mass and half-mass radius as in the Baumgardt catalog.  We then integrate the evolution of these particles forward in time to the present day, by taking into account that each particle feels the gravitational potential of the Galaxy, that of its host cluster, and that of all the other clusters in the Galaxy. Note that we do not take into account self-gravity among particles. The particles experience the gravitational field yet do not contribute to it. The equation of motion of a generic particle among the 100,000 which populate Palomar~5 is thus: 
        \begin{equation}
          \ddot{\vec{r}}_p = -\nabla \Phi(R_p,z_p) + \left.\sum_{j}^{N_{GC}} \frac{Gm_j}{\left(|\vec{r}_j(t) - \vec{r}_p|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j(t)- \vec{r}_p\right),
          \end{equation} \label{eq:equation_of_motion_particle} where the index $p$ represents one of the 100,000 particles of interest, $\vec{r_p}$ being its position, and $j$ indexes over the globular clusters as in Eq.~\ref{eq:GCNBody}. Note in that Eq.~\ref{eq:equation_of_motion_particle} the positions of the globular clusters are time-dependent since they are being loaded during this step and not computed, unlike Eq.~\ref{eq:GCNBody}.

@@

    The integration was conducted using a leapfrog integrator, chosen for its ability to preserve phase-space volume and conserve the Hamiltonian with each integration step. This method is preferable to, for instance, a Runge-Kutta scheme, which can introduce non-physical and significant numerical errors in systems that require long-term stability and energy conservation. One drawback of the leapfrog integrator is that it requires a uniform time-step across the entire computation, leading to unnecessary computations for a particle after it escapes form the host cluster. However, energy conservation and phase-space volume preservation are paramount when modeling stellar streams. The time step was therefore set to be small enough to conserve energy for the most interior particles within the cluster---ensuring that a higher mass loss did not arise from numerical error. For these simulations, a time-step of 10,000 years was selected to maintain adequate energy conservation of an median variation of $10^{-12} \frac{\Delta E}{E_0}$.

    We also checked the \textit{reverse integrability} of the globular cluster system. By reverse integrability, we mean the integrator's capability to track the cluster backward in time and then re-integrate it forward along the same trajectory. The drift in the final position after forward integration, compared to the initial position from the backward integration, was consistently at least two orders of magnitude smaller than the Plummer scale radius used for Palomar~5. This high precision ensures that no fictitious evil numerical forces influenced the system, preventing any artificial mass loss or retention of star particles.

@@

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{decomposition-monte-carlo-009-with-3-gaps-domidpoint-shift.png}
      \caption{Density maps of Palomar~5's stream in the tail coordinate system. The colorscale represents normalized particle counts (total: 100,000). The top panel shows the \texttt{full} simulation with three gaps on the stream's right-hand side. The next three panels depict simulations with identical initial conditions but exclude the gravitational influence of all clusters except those forming a given gap. The \texttt{Reference} simulation omits other globular clusters. The bottom panel compares Palomar5's observed stream length to the \texttt{Reference} simulation, using the same Monte Carlo realization as Fig.~\ref{fig:stream_on_sky} and \texttt{Sampling~009} of Fig.~\ref{fig:gallery2}.}
      \label{fig:decomposition}
    \end{figure*}

@@

    To make a more quantitative comparison between the \texttt{reference} and \texttt{full} simulations, we work in the tail coordinate system, in which the stream is aligned with the cluster's orbit. The definition of this coordinate system is based on the work of \citet{2004AJ....127.2753D} and is shown in Fig.~\ref{fig:TailCoordinates}. Briefly, in this system the $x_{\textrm{tail}}$ coordinate represents the position of a particle along the orbit relative to the globular cluster. Positive values of $x_{\textrm{tail}}$ are ahead of the cluster while negative $x_{\textrm{tail}}$ is behind the cluster. The $y_{\textrm{tail}}$ coordinate measures the particle's distance within the orbital plane, where positive values indicate that the particle is farther from the Galactic center and negative values indicate that it is closer. Fig.~\ref{fig:decomposition} shows a comparison between one of the 50 realizations of Palomar~5 stream taking into account the gravitational interactions with all other globular clusters in the Galaxy (top panel) and omitting them (bottom panel). This comparison clearly shows the presence of two wide ($\sim$ 100~pc and $\sim$ 1~kpc) gaps in the leading tail and of a more subtle under-density at $x_{\textrm{tail}}\sim 5$~kpc (we refer the reader to Appendix~\ref{sec:gap_detection} for a detailed description of the under-densities and gaps detection method).

    To find out which globular clusters were responsible of creating these gaps, and when close passages took place, we estimated the gravitational acceleration along the orbit of Palomar~5 and represented it in the ($t$, $\tau$) space. $t$ is the simulation time and $\tau$ is indicates how long it will take for Palomar~5 to reach, or how  long ago it passed, a given point on its orbit. The use of $\tau$ is advantageous because the growth of the stream is approximately linear in $\tau$. On the other hand, in physical space streams on eccentric orbits indeed grow in time yet the growth rate is modulated with periodic expansion and contraction depending on the orbital phase (see the top panel of Fig.~5 from \citet{2016MNRAS.457.3817S} for an example). Adopting this time space, and reporting the gravitational acceleration along Palomar~5 orbit in this space, the identification of the globular clusters which produced the perturbation, and the time at which it occurred becomes straight forward. We refer the reader to Appendix~\ref{sec:Perturber_Identification} for an in-depth description of the procedure. In this way, we could identify that the clusters responsible for creating gaps in the simulation of Palomar~5 stream reported in Fig.~\ref{fig:decomposition} are NGC~2808, NGC~7078 and NGC~104, and that their close passages occurred 200~Myr, -1.9 and -2.1~Gyr ago, respectively.

@@

    This computation allows us to analyze the distribution of gap creation rates across all simulations. Notice that since the gap creation rate \textit{adds in parallel}, naturally gaps that occur at earlier times when the stream was shorter are weighted higher than those that occur when the stream is longer. The results are shown in Fig.~\ref{fig:gapcreationrate}. This is roughly consistent with a simple estimate of the average gap creation rate: with 73 gaps over 5 Gyr of integration time for a stream about 20 kpc in length, the naive estimate is approximately 0.015~km~s$^{-1}$~kpc$^{-2} $ (which is roughly equivalent to $0.015~\rm{Gyr^{-1}kpc^{-1}}$). This is about double the weighted mean gap creation rate of 0.009~km~s$^{-1}$~kpc$^{-2}$. The naive estimate is higher because it does not account for the growth of the stream over time, unlike Eq.\ref{eq:gap_creation_rate}.

@@

    While Fig.~\ref{fig:impact_geometry_statistics} demonstrates that mass, relative velocity, or impact parameter alone cannot predict gap formation, one interesting result emerges: no gaps are created from interactions with impact parameters greater than 300~pc. The stream widths are roughly 200~pc. This finding is even more evident when examining the $b$-$M$ plane. A series of perturbers at roughly $\sim7 \times 10^5 M_\odot$ highlights NGC~2808's flybys, where all encounters with impact parameters under 200~pc result in gaps, while those beyond this distance do not. In other words, massive globulars with masses greater than $10^6 \rm{M}_\odot$ are not effective in producing gaps if they pass close to Palomar~5 stream with an impact parameter $b > 200$~pc. It is interesting to note that for values of the impact parameter below this threshold, even fast encounters ($\delta v > 300$~km/s) can produce gaps.

@@

    Once all the key impact parameters estimated, we can use Eq~\ref{eq:change_in_momentum} and calculate $\Delta p$, the change in momentum (per unit mass) imparted by a cluster flyby on Palomar~5 stream. The result  is shown in Fig.~\ref{fig:deltap}, where the distribution of imparted in change in momentum is shown for all impacts which produce a gap, and compared to those that do not produce one. On average, encounters which lead to gap creations impart a change in momentum on stream particles which is a factor of 10 higher than that of encounters which do not form gaps (but with some overlap in the low-velocity tail). Interestingly, changes in momentum which lead to gap creations extend over a large range in velocities. There is a factor of about 100 between the smallest and largest changes with NGC~2808 (orange color in the histogram) imparting changes in the velocity of stream particles which redistribute over the whole range of $\Delta p$.

@@

    Another possible explanation is the different thickness of the simulated tails at different distances from the center of the cluster. At distances of less than about 3~kpc from Palomar~5, the tails are very thin, with a typical thickness being less than 100 pc. In order to cause gaps in these regions, the clusters would have to pass close to the stream, with impact parameters similar to the thickness itself. Note that the thinness of the tails at these distances is probably a direct consequence of the initial parameters we chose for the simulation. We have assumed for Palomar~5, 5 Gyr years ago, the same internal parameters (mass and size) that the cluster has today. With such parameters, after 5 Gyr of evolution, our system has lost most of its mass, and therefore the part of the tails closest to the cluster itself, whose density depends mainly on the most recent mass loss \citep[see, for example, Fig.~A.3 in][]{2012A&A...546L...7M}, is necessarily very thin because the simulated cluster has essentially no more mass to lose. This last point has also a consequence in the gap creation rate, which we derived at the end of Sect.~\ref{sect:geometry}: with only 70\% of the tail  ($\sim$ 14~kpc over 20~kpc, excluding the innermost $\pm 3$~kpc from the cluster center) really suitable for forming gaps, the gap creation rates are about 50$\%$ higher than those estimated in the aforementioned section, where values have been derived taking into account the full tail extent. This gap creation rate would be still not high enough to reproduce the  number of gaps in Palomar~5 tails, as reported by \citet{2012ApJ...760...75C}. In their study, they indeed suggested the presence of  five gaps with a 99~\% detection confidence, leading to a gap creation rate of 0.17~Gyr$^{-1}$~kpc$^{-1}$. If true, this rate would be too high to be explained by Galactic globular clusters only.

@@

  Our simulations produce a stream for Palomar~5 that is longer than the currently observed extent. In simulations, stream detections are straightforward because we can use reference runs to clearly separate stream particles from the field and compare against a known "true" structure. In contrast, observational data are inherently more challenging due to magnitude limits, contamination from field stars, and the lack of a ground truth for comparison. These factors suggest that the number of gaps identified in this study could represent an upper limit.

>>>>

\institute{
  Paris Observatory. Paris Sciences et Lettres\\
  \email{salvatore.ferrone@obspm.fr}
  \and
    Dipartimento di Fisica, Universit\`a di Roma ``La Sapienza'', Piazza Aldo Moro
  \and
    Dipartimento di Fisica e Astronomia. ``Galileo Gallilei'' Università di Padova, Vicolo dell'Osservatorio 3 Padova 35122, Italy.
    \and
      Universit\'e de Strasbourg, CNRS, Observatoire astronomique de Strasbourg, UMR 7550, F-67000 Strasbourg, France
    \and
      Leibniz-Institut für Astrophysik Potsdam (AIP), An der Sternwarte 16, 14482 Potsdam, Germany
    \and
      Exeter
    % \thanks{}
  }

@@

  One of the interests in studying stellar streams is that they can constrain the gravitational field of their host galaxies, particularly the Milky Way. Compared to the measurement of the HI rotation curve, stellar streams offer the opportunity to investigate the potential of the host galaxy over a wide range of distances, reaching the outermost regions of the halo. For example, \citet{2011MNRAS.417..198V} demonstrated how stellar streams can infer the mass and scale parameters of dark matter halos using various amounts of observational data, from basic right ascension and declination to full six-dimensional phase space information. \citet{2018ApJ...867..101B} reviewed this concept from an information-theoretic point of view, identifying which orbits and configurations of stellar streams yield the most information about the galactic potential. However, the use of single streams for constraining the potential led to some ambiguous and not converging results over time. For example, \citet{2010ApJ...718.1128L} made use of the Sagittarius stream to infer that the dark matter halo of our Galaxy has a triaxial shape \citep[but see also][]{2004MNRAS.351..643H, 2005ApJ...619..800J, 2005ApJ...619..807L}; \citet{2016ApJ...833...31B} concluded that the dark matter halo of our Galaxy is nearly spherical at the distances of the Palomar~5 and GD-1 streams. The two contrasting results could also show that at one distance is the halo is triaxial and at another spherical--highlighting the need for streams at different distances to map the halo shape. Recently, \citet{2024ApJ...967...89I} constructed the most accurate model of the Milky Way to date by applying a Monte Carlo Markov Chain (MCMC) fitting procedure. This method identified the set of potential parameters in an axisymmetric model of the Milky Way that best reproduces all observed stellar streams. This work represents a seminal achievement and a significant milestone in the field.

@@

\section{Methods}
  The most accurate model of the Palomar~5 stream would involve full modeling of the internal dynamics of the cluster. This would mean computing $N$-Body interactions with a $\mathcal{O}(N^2)$ computation time, stellar evolution, super novae, an initial mass distirbution, treatment of binary star systems, etc. \citep[for such an example][see,]{2021NatAs...5..957G, 2016MNRAS.458.1450W}. Instead, We opt for solving the restricted-three body problem or particle-test method as we did for \citet{2023A&A...673A..44F}, which we present here for completeness.

@@

    Lastly, we select an integration time of 5~Gyr as a compromise between maximizing interaction statistics and adhering to the assumption that the galaxy can be modeled as a time-independent, constant mass distribution. \citet{2023A&A...673A.152I} analyzed the orbits of the Galactic cluster population using the same initial conditions as in this work, within five live Milky Way-like potentials from IllustrisTNG \citep{2018MNRAS.473.4077P}. They found that in all sampled potentials, orbital changes remain minimal over 5~Gyr, becoming significant only at earlier look-back times when the host galaxy had significantly less mass or was undergoing a merger event.

    \subsubsection*{ \texttt{Full} simulations}
    There is a primary methodological departure from \citet{2023A&A...673A..44F}.  In that paper, globular clusters evolved under the gravitational effect of the Galaxy only, while now we include also the effect of all the other Galactic globular clusters, by taking into account the direct $N$-body interactions among them. To quantify these interactions, all clusters are represented by Plummer spheres, each with mass and half-mass radius as reported in the Baumgardt catalog \citep{2021MNRAS.505.5957B}. For the remainder of this paper, we refer to these simulations, where the globular clusters interactions are taken into account as \texttt{full} simulations. \\
    For these simulations, we proceed into two steps:
      \begin{enumerate}
          \item First, starting from the Galactocentric positions and velocities of all 165 Galactic globular clusters, we integrate their orbits back in time for 5~Gyr under the influence of the Galaxy itself and their mutual influence. In the backward integration, the system of equations of motion for the globular clusters are thus: 
          \begin{equation}
          \ddot{\vec{r}}_i = -\nabla \Phi(R_i,z_i) + \left.\sum_{j\neq i}^{N_{GC}} \frac{Gm_j}{\left(|\vec{r}_j - \vec{r}_i|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j - \vec{r}_i\right),
          \end{equation}\label{eq:GCNBody} where $\vec{r}$ indicates the galactocentric position vector, the index $i$ indicates the globular cluster of interest with $R_i,z_i$ being its galactocentric positions in cylindrical coordinates; the index $j$ indicates the other globular clusters that are summed over. $N_{GC}$ is the total number of globular clusters which in this study is 165, $m_j$ is the mass of the j-th cluster in the sample, $b_j$ is its Plummer scale radius, and $\vec{r_j}$ is its galactocentric position. $\Phi$ represents the Galactic smooth potential \citep[][Model~II, in the present case]{2017A&A...598A..66P}. Note that the masses and sizes of the globular clusters are kept constant in these simulations and not allowed to vary with time. This equally implies that we do not take into account the internal evolution of the globular clusters in these models. We discuss the implications of the modeling limitations in the Discussion.

        \item Once the positions and velocities of the entire globular cluster system, as to 5~Gyr ago, are found, we sample Palomar~5 with  100,000 particles which follow a Plummer distribution of mass and half-mass radius as in the Baumgardt catalog.  We then integrate the evolution of these particles forward in time to the present day, by taking into account that each particle feels the gravitational potential of the Galaxy, that of its host cluster, and that of all the other clusters in the Galaxy. Note that we do not take into account self-gravity among particles. The particles experience the gravitational field yet do not contribute to it, which is a common assumption made in Galactic dynamics since the mass of an individual star is negligible compared to the mass of the larger dynamical system. The equation of motion of a generic particle among the 100,000 which populate Palomar~5 is thus: 
        \begin{equation}
          \ddot{\vec{r}}_p = -\nabla \Phi(R_p,z_p) + \left.\sum_{j}^{N_{GC}} \frac{Gm_j}{\left(|\vec{r}_j(t) - \vec{r}_p|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j(t)- \vec{r}_p\right),
          \end{equation} \label{eq:equation_of_motion_particle} where the index $p$ represents one of the 100,000 particles of interest, $\vec{r_p}$ being its position, and $j$ indexes over the globular clusters as in Eq.~\ref{eq:GCNBody}. Note in that Eq.~\ref{eq:equation_of_motion_particle} the positions of the globular clusters are time-dependent since they are being loaded during this step and not computed, unlike Eq.~\ref{eq:GCNBody}.

@@

    The integration was conducted using a leapfrog integrator, chosen for its ability to preserve phase-space volume and conserve the Hamiltonian with each integration step. This method is preferable to, for instance, a Runge-Kutta scheme, which can introduce non-physical and significant numerical errors in systems that require long-term stability and energy conservation. One drawback of the leapfrog integrator is that it requires a uniform time-step across the entire computation, leading to unnecessary computations for a particle after it escapes form the host cluster. However, energy conservation and phase-space volume preservation are paramount when modeling stellar streams. The time step was therefore set to be small enough to conserve energy for the most interior particles within the cluster---ensuring that a higher mass loss did not arise from numerical error. For these simulations, a time-step of 10,000 years was selected to maintain adequate energy conservation of an median variation of $10^{-12} \frac{\Delta E}{E_0}$, where $E_0$ is a particle's initial energy, and $\Delta E$ is the difference between its final and initial energy.

    We also checked the \textit{reverse integrability} of the globular cluster system for the \texttt{reference} simulations. By reverse integrability, we mean the integrator's capability to track the cluster backward in time and then re-integrate it forward along the same trajectory. Integrating point masses in a static axis-symmetric potential conserves $L_z$ and $E$, which create regular periodic and non-chaotic orbits. Therefore, any drift would arise from purely numerical error. We selected a time stamp for which the drift in the final position after forward integration, compared to the initial position from the backward integration, was consistently at least two orders of magnitude smaller than the Plummer scale radius used for Palomar~5. This high precision ensures that no fictitious numerical forces influenced the system, preventing any artificial mass loss or retention of star particles.

@@

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{decomposition-monte-carlo-009-with-3-gaps-domidpoint-shift.png}
      \caption{Density maps of Palomar~5's stream in the tail coordinate system. The colorscale represents normalized particle counts (total: 100,000). The top panel shows the \texttt{full} simulation with three gaps on the stream's right-hand side. The next three panels depict simulations with identical initial conditions but exclude the gravitational influence of all clusters except those forming a given gap. The \texttt{Reference} simulation omits the influence from other globular clusters. The bottom panel compares Palomar~5's observed stream length to the \texttt{Reference} simulation, using the same Monte Carlo realization as Fig.~\ref{fig:stream_on_sky} and \texttt{Sampling~009} of Fig.~\ref{fig:gallery2}.}
      \label{fig:decomposition}
    \end{figure*}

@@

    To make a more quantitative comparison between the \texttt{reference} and \texttt{full} simulations, we work in the tail coordinate system, in which the stream is aligned with the cluster's orbit. The definition of this coordinate system is based on the work of \citet{2004AJ....127.2753D} and is shown in Fig.~\ref{fig:TailCoordinates}. Briefly, in this system the $x_{\textrm{tail}}$ coordinate represents the position of a particle along the orbit relative to the globular cluster. Positive values of $x_{\textrm{tail}}$ are ahead of the cluster while negative $x_{\textrm{tail}}$ is behind the cluster. The $y_{\textrm{tail}}$ coordinate measures the particle's distance within the orbital plane, where positive values indicate that the particle is farther from the Galactic center and negative values indicate that it is closer. Fig.~\ref{fig:decomposition} shows a comparison between one of the 50 realizations of Palomar~5 stream taking into account the gravitational interactions with all other globular clusters in the Galaxy (top panel) and omitting them (bottom panel). This comparison clearly shows the presence of two wide ($\sim$100~pc and $\sim$1~kpc) gaps in the leading tail and of a more subtle under-density at $x_{\textrm{tail}}\sim 5$~kpc (we refer the reader to Appendix~\ref{sec:gap_detection} for a detailed description of the under-densities and gaps detection method).

    To find out which globular clusters were responsible of creating these gaps, and when close passages took place, we estimated the gravitational acceleration along the orbit of Palomar~5 and represented it in the ($t$, $\tau$) space. $t$ is the simulation time and $\tau$ is indicates how long it will take for Palomar~5 to reach, or how  long ago it passed, a given point on its orbit. The use of $\tau$ is advantageous because the growth of the stream is approximately linear in $\tau$. On the other hand, in physical space streams on eccentric orbits indeed grow in time yet the growth rate is modulated with periodic expansion and contraction depending on the orbital phase ( \citep[see the top panel of Fig.~5.][for an example]{2016MNRAS.457.3817S}). Adopting this time space, and reporting the gravitational acceleration along Palomar~5 orbit in this space, the identification of the globular clusters which produced the perturbation, and the time at which it occurred becomes straight forward. We refer the reader to Appendix~\ref{sec:Perturber_Identification} for an in-depth description of the procedure. In this way, we could identify that the clusters responsible for creating gaps in the simulation of Palomar~5 stream reported in Fig.~\ref{fig:decomposition} are NGC~2808, NGC~7078 and NGC~104, and that their close passages occurred 200~Myr, -1.9 and -2.1~Gyr ago, respectively.

@@

    This computation allows us to analyze the distribution of gap creation rates across all simulations. Notice that since the gap creation rate \textit{adds in parallel}, naturally gaps that occur at earlier times when the stream was shorter are weighted higher than those that occur when the stream is longer. The results are shown in Fig.~\ref{fig:gapcreationrate}. This is roughly consistent with a simple estimate of the average gap creation rate: with 73 gaps over 5~Gyr of integration time for a stream about 20 kpc in length, the naive estimate is approximately 0.015~km~s$^{-1}$~kpc$^{-2} $ (which is roughly equivalent to $0.015~\rm{Gyr^{-1}kpc^{-1}}$). This is about double the weighted mean gap creation rate of 0.009~km~s$^{-1}$~kpc$^{-2}$. The naive estimate is higher because it does not account for the growth of the stream over time, unlike Eq.\ref{eq:gap_creation_rate}.

@@

    While Fig.~\ref{fig:impact_geometry_statistics} demonstrates that mass, relative velocity, or impact parameter alone cannot predict gap formation, one interesting result emerges: no gaps are created from interactions with impact parameters greater than 300~pc. The stream widths are roughly 200~pc, as seen in Appendix~\ref{sec:gallery_of_gaps}. This finding is even more evident when examining the $b$-$M$ plane. A series of perturbers at roughly $\sim8 \times 10^5 M_\odot$ highlights NGC~2808's flybys, where all encounters with impact parameters under 200~pc result in gaps, while those beyond this distance do not. In other words, even the most massive globulars with masses greater than about $10^6 \rm{M}_\odot$ cannot cause gaps if there impact parameters are greater than about $300$~pc. It is interesting to note that for values of the impact parameter below this threshold, even fast encounters ($\delta v > 300$~km/s) can produce gaps. Perhaps, this is not surprising since the range of possible relative velocities is much less than mass and impact parameter which vary two and three orders, respectively while the relative velocities only vary by about a factor of three.

@@

    Once all the key impact parameters estimated, we can use Eq.~\ref{eq:change_in_momentum} and calculate $\Delta p$, the change in momentum (per unit mass) imparted by a cluster flyby on Palomar~5 stream. The result  is shown in Fig.~\ref{fig:deltap}, where the distribution of imparted in change in momentum is shown for all impacts which produce a gap, and compared to those that do not produce one. On average, encounters which lead to gap creations impart a change in momentum on stream particles which is a factor of 10 higher than that of encounters which do not form gaps (but with some overlap in the low-velocity tail). Interestingly, changes in momentum which lead to gap creations extend over a large range in velocities. There is a factor of about 100 between the smallest and largest changes with NGC~2808 (orange color in the histogram) imparting changes in the velocity of stream particles which redistribute over the whole range of $\Delta p$.

@@

    Another possible explanation is the different thickness of the simulated tails at different distances from the center of the cluster. At distances of less than about 3~kpc from Palomar~5, the tails are very thin, with a typical thickness being less than 100~pc. In order to cause gaps in these regions, the clusters would have to pass close to the stream, with impact parameters similar to the thickness itself. Note that the thinness of the tails at these distances is probably a direct consequence of the initial parameters we chose for the simulation. We have assumed for Palomar~5, 5~Gyr years ago, the same internal parameters (mass and size) that the cluster has today. With such parameters, after 5~Gyr of evolution, our system has lost most of its mass, and therefore the part of the tails closest to the cluster itself, whose density depends mainly on the most recent mass loss \citep[see, for example, Fig.~A.3 in][]{2012A&A...546L...7M}, is necessarily very thin because the simulated cluster has essentially no more mass to lose. This last point has also a consequence in the gap creation rate, which we derived at the end of Sect.~\ref{sect:geometry}: with only 70\% of the tail  ($\sim$ 14~kpc over 20~kpc, excluding the innermost $\pm 3$~kpc from the cluster center) really suitable for forming gaps, the gap creation rates are about 50$\%$ higher than those estimated in the aforementioned section, where values have been derived taking into account the full tail extent. This gap creation rate would be still not high enough to reproduce the  number of gaps in Palomar~5 tails, as reported by \citet{2012ApJ...760...75C}. In their study, they indeed suggested the presence of  five gaps with a 99~\% detection confidence, leading to a gap creation rate of 0.17~Gyr$^{-1}$~kpc$^{-1}$. If true, this rate would be too high to be explained by Galactic globular clusters only.

@@

  Our simulations produce a stream for Palomar~5 that is longer than the currently observed extent. In simulations, stream detections are straightforward because we can use reference runs to clearly separate stream particles from the field and compare against a known ``true'' structure. In contrast, observational data are inherently more challenging due to magnitude limits, contamination from field stars, and the lack of a ground truth for comparison. These factors suggest that the number of gaps identified in this study could represent an upper limit.
